WORKING TITLE: The adaptive value of green nest material across birds: A systematic review and meta-analysis

### 2025-06-26 manuscript_GNM.qmd
Introduction update.. continued.. no built

>>>>

# Introduction

Nest construction is a fundamental aspect of the bird breeding cycle,

Nests provide structural support and protection for eggs, parents, and offspring against predators and environmental pressures, and are thus key to reproductive success and survival [@mainwaring2014; @hansell2000]. While most birds use organic plant materials such as twigs, branches, grass, and moss, some incorporate inorganic items such as stones, feathers, snake skin and even anthropogenic materials [@sergio2011; @sheard2024]. These non-structural additions suggest that nests are more than just physical shelters; they can be complex tools shaped by both natural and sexual selection for multiple functions. Indeed, nests can play a key role in sexual display, with some features such as size acting as sexually selected signals of parental quality in many bird species (XX), and even being used interspecifically by brood parasites (XX). Nest building behaviour is increasingly recognized as an expression of extended phenotype (reviewed by @moreno2012), and phylogeny has been shown to be an important factor explaining both mean and within-species variation in nest morphology [@perez2023]. Yet, despite hundreds of scientific articles studying nests and nesting behaviour in birds being published each year, much remains unknown.

An intriguing nesting behaviour observed in some bird species is the addition of fresh, often aromatic, plant material to the nest. This green nest material (*GNM*) includes leaves, sprigs, herbs and branches often added from the onset of nest building to the start of egg laying, but in some species, such as the Blue Tits, also after primary nest building is completed. during or even after primary nest construction has completed. Importantly, contrary to the structural nest materials or components, this GNM are thought to be selected from a small, non-random subset of local flora [@clark1985; @garrido-bautista2023], often characterized by high levels of aromatic volatile compounds (commonly used by Passerines: @mennerat2009) or secondary metabolites such as oils and resins (commonly used by raptors: @ontiveros2007). The costs associated with gathering these materials beyond the territory and maintaining fresh greenery through the breeding season imply adaptive significance of the GNM. <!--# This information is important, but it’s not entirely clear why this behaviour is considered costly. Also, we haven’t yet mentioned that birds maintain this material fresh throughout the nesting period, which is key to arguing that it requires extra effort. We need to explain this point more clearly and provide evidence for the associated costs. Are there any studies that directly assess these costs? For instance, does this paper—https://journals.sagepub.com/doi/10.3184/175815512X13528994072997—address it? Please add this information, and I’ll be happy to help revise the text further. -->

Several hypotheses have been proposed to explain the adaptive function of GNM [@dubiec2013; @scott-baumann2015]. First, the courtship hypothesis suggests that mostly males use GNM signals to attract females or advertise nest site occupancy, particularly in cavity-nesting species since visual cues inside the nest are limited. This hypothesis traces back to the early natural history observations by @kessel1957, who noted that male European Starlings (*Sturnus vulgaris*) carried fresh plant material to their nests while singing and displaying to females. Subsequent studies in borh European and Spotless Starlings (*Sturnus unicolor*) supported this view that GNM may serve as a sexually selected signal of male quality by showing that (i) unpaired males conspicuously deliver GNM during courtship, stopping right after the onset of egg laying [@brouwer2004; @veiga2006], and seemingly increasing this behaviour in the presence of potential mates [@eens1993], and (ii) females often prefer nests with fresh GNM. The courtship hypothesis was later extended to explain GNM addition solely by females in species like Blue Tits (*Cyanistes caeruleus*), therefore suggesting that GNM might serve as a sexually selected of female quality [@tomás2013].

Second and the most widely tested one, is the nest protection hypothesis, which proposes that the volatile compounds in GNM reduce the amount of parasites or pathogens in the nests, enhancing offspring survival. This hypothesis traces back to a comparative study in raptors (Falconiformes) that proposed it and provided indirect evidence for it by comparing the nest reuse, diet composition and GNM presence \[@wimberger1984\]. Soon after, a preference for GNM containing insect-repellent and antimicrobial volatile was experimentally confirmed in European Starlings, therefore confirming two of the three key criteria for evaluating the nest protection hypothesis (i.e., selective plant use, bioactive chemical properties and demonstrable adaptive benefits; @clark1988). Observational and experimental studies on raptors \[@ontiveros2007; @dykstra2009\] and passerines [@dawson2004; @lafuma2001; @mennerat2009b; @quiroga2012] have shown mixed results, and in some cases, GNM has been shown to improve offspring condition without reducing the amount of parasites in the nests [@gwinner2000], which led to the development of drug hypothesis. The drug hypothesis suggests that the volatiles present in GNM directly enhance offspring immune function and/or health independent of parasite or pathogen control [@gwinner2000; @mennerat2009a]. Thus, while both hypotheses suggest adaptive benefit of GNM by enhancing offspring survival, the drug hypothesis emphasizes a pharmacological mechanism, whereas the nest protection hypothesis emphasizes parasite and pathogen repellency. Importantly, these two hypotheses are not mutually exclusive unless the improvement in offspring health can be proved to be unrelated to a reduction in parasite or pathogen load.

Last, several additional adaptive hypotheses or functions of GNM that do not refer to its aromatic properties have been proposed. These include benefits such as enhancing the crypticity of the nest [@skutch1957], providing insulation or sun protection for eggs and offspring [@mertens1977], or simply fulfilling a structural function [@lyons1987]. However, these alternative functional roles have received little support or attention. In all, despite decades of research, abundant observational and experimental studies, and two crucial narrative reviews [@dubiec2013; @scott-baumann2015], we still do not know the overall adaptive significance of GNM, its generality across bird species, and the relative importance of each of the suggested mechanisms. Given that nest building is widespread across taxa ---including but not limited to birds--- and is increasingly recognized as an expression of the extended phenotype, a quantitative synthesis of this field is both timely and potentially far-reaching in its implications.

To address this gap, we conducted a pre-registered systematic review and meta-analysis of experiments testing the effect of GNM on various fitness proxies across birds. Our meta-analysis explicitly evaluates the overall evidence for the adaptive significance of GNM, as well as to assesses the relative importance of the main suggested hypotheses: the courtship hypothesis and the parental care hypothesis (i.e. combining both the nest protection and drug hypotheses due to their non-exclusivity). In addition, we explore the observed context-dependency (i.e., high heterogeneity among effect sizes) via meta-regressions. Specifically, we tested whether the effect of GNM varied depending on the type of parasite (arthropods vs. microorganisms), the timing of GNM addition (before hatching, after hatching, or continuously), the experimental design (aromatic vs. non-aromatic vs. no added material), the focal species, and the type of fitness proxy investigated (e.g., physiology, morphology, reproduction). Our meta-analysis of experimental studies is the first quantitative synthesis testing the causal relationship between GNM and fitness across birds.
